---
title: "Assignment 3"
subtitle: "Unsupervised Learning"
date: last-modified
date-format: "D MMMM YYYY"
authors:
  - name: Raphaela Azar
  - name: Shvet Maharaj
format: 
  pdf:
    pdf-engine: pdflatex
    papersize: "a4paper"
    include-in-header: 
      text: |
        \newcommand{\B}[1]{\boldsymbol{#1}}
        \newcommand{\+}{\phantom{-}}
        \newcommand{\op}[1]{\operatorname{#1}}
execute:
  echo: false
  warning: false
  freeze: auto
---

# Introduction

BRIEF INTRODUCTON TO OUR PROJECT

# Dataset

We are utilising the `` dataset. Let us explore this dataset.

```{r}
#| label: libraries
#| include: false
library(tidyverse, attach.required = T)
library(future)
library(doFuture)
library(progressr)
library(parallelly)
```

```{r}
#| label: parallel-computation-setup
options(parallelly.availableCores.omit = 1)
if(supportsMulticore()) # multicore = mac, linux, etc. (outside of RStudio)
  plan(multicore) else plan(multisession) #multisession = windows and other
on.exit(plan(sequential)) # to ensure that the workers are correctly shut down
```

```{r}
#| label: theming
maincolour <- "darkblue"
# set ggplot defaults
base_theme <- theme_minimal() + theme(aspect.ratio=1)
theme_set(base_theme)
```

```{r}
#| label: data-cleaning
raw_dat <- read.table("STA4026_Assignment_Clustering.txt") |> as_tibble()

raw_dat |> summarise(n = n(),
                     distinct    = n_distinct(pick(V1,V2)),
                     distinct_V1 = n_distinct(V1), 
                     distinct_V2 = n_distinct(V2),
                     any_na      = anyNA(pick(V1,V2), recursive = T))
# Result: 
# No duplicate points. 
# Few points sharing one X or Y value
# No NA values.
# No need to remove values.
dat <- raw_dat |> mutate(x = V1, y = V2, .keep='unused')
# standardise???
```

The data has been cleaned through removing the entries in which any value is missing.
Duplicate points have been removed.
But! None were duplicates, and none were missing. So the data remains as is.

Neither variable has been named, so for ease of discussion, the first variable will be referred to as $\B x$, and the second variable, $\B y$.

```{r}
#| label: fig-data-vis
dat |> ggplot(aes(x=x,y=y)) + geom_point(alpha=0.3,size=0.5,colour=maincolour)
```

# Cross-Validation

# Cluster Analysis

# Conclusion